---
title: Lecture 01b -- Revision
author: Januar Harianto
# format: soles-revealjs
format: revealjs
fontsize: 16pt
# embed-resources: true
execute:
  cache: true
---

```{r}
#| label: setup
#| include: false
#| warning: false
#| message: false

# Load required packages
pacman::p_load(tidyverse, patchwork)

# Set random seed for reproducibility
set.seed(909)

# Set default ggplot theme
theme_set(theme_minimal())
```

# Refresher

## Things to recall

- Samples, populations and statistical inference
- Probability distributions
- Parameter estimation
  - central tendency
  - spread or variability
- Sampling distributions (mean)
- Standard error (mean)
- Confidence intervals (mean)
- Controlled experiments vs observational studies

# Samples, populations and statistical inference

## Populations and Samples

### Populations
- **All** the possible units and their associated observations of interest
- Scientists are often interested in making *inferences* about populations, but measuring every unit is impractical

### Samples
- A collection of observations from any population is a sample, and the number of observations in it is the **sample size**
- We assume samples that we collect can be used to make inferences about the population
- **NEW**: Samples need to be *representative* of the population

::: fragment
The basic way of generating a sample is **random sampling**, where every observation has the same probability of being included in the sample. We will go in-depth on this next week.
:::

## Statistics vs Parameters

- Characteristics of the **population** are called *parameters* (e.g. population mean or population regression slope)
- Characteristics of the **sample** are called *statistics* (e.g. sample mean or sample regression slope) -- they are used to estimate the population parameters
- Statistics are what we use to help use understand the population
- Formal statistical methods can help us make inferences about the population based on the sample -- statistical inference
- *Not all statistical techniques are inferential, but many are*



## Sample Data

Sample data are usually collected as **variables**, which are the characteristics we measure or record from each object.

::: fragment

Variables can be:

### Categorical Variables
- **Nominal**: categories without a natural order (e.g. colors, names)
- **Ordinal**: categories with a natural order (e.g. ratings, rankings)

### Numerical Variables
- **Continuous**: can take any value within a range (e.g. height, weight)
- **Discrete**: can take only specific values (e.g. counts, presence/absence)
:::

## YOU decide on what a variable represents

A numerical, continuous variable can be treated as a categorical variable if you decide to categorise it.

### Examples
::: incremental
- **height (in cm)** -- a numerical, **continuous** variable, can be treated as a categorical variable if you group it into categories (short, medium, tall)
- **age (in years)** -- a numerical, **discrete** variable, can be treated as a continuous variable (if we allow for certain *issues*)
- **treatment (A, B, C)** -- a categorical variable, can be treated as a numerical variable if we assign numbers to the treatments (1, 2, 3) and assume they are **ordered** e.g. effect of 1 < 2 < 3 -- *the basis of non-parametric tests*
:::

# Distribution of data
## Types of probability distributions
By now, you should be familiar with these distributions and their properties

::: incremental
- **Normal Distribution**: Bell-shaped curve, symmetric around the mean. **Data is continuous**
- **Binomial Distribution**: Models success/failure outcomes in a fixed number of trials. **Data is discrete**
- **Poisson Distribution**: Models count data when events occur at a constant rate. **Data is discrete**
:::
::: fragment
Knowing the distribution of your data is important for choosing the right statistical model -- although it is not always necessary.
:::

## 
```{r}
library(tidyverse)

set.seed(908)
normal_data <- data.frame(x = rnorm(10000, mean = 0, sd = 1))
binomial_data <- data.frame(x = rbinom(10000, size = 10, prob = 0.5))
poisson_data <- data.frame(x = rpois(500, lambda = 3))

# normal
p1 <- ggplot(normal_data, aes(x = x)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue", color = "black") +
  geom_density(color = "red") +
  ggtitle("Normal Distribution") +
  theme_minimal()

# binomial
p2 <- ggplot(binomial_data, aes(x = x)) +
  geom_bar(fill = "lightgreen", color = "black") +
  ggtitle("Binomial Distribution") +
  theme_minimal()

# poisson
p3 <- ggplot(poisson_data, aes(x = x)) +
  geom_bar(fill = "lightpink", color = "black") +
  ggtitle("Poisson Distribution") +
  theme_minimal()

# Arrange plots
library(patchwork)
p1 | p2 / p3
```

# Parameter estimation

## Measures of Central Tendency

### Mean
::: incremental
- The arithmetic average of all values in a dataset
- Sum of all values divided by number of observations
- Sensitive to extreme values (outliers)
- Formula: $\bar{x} = \frac{1}{n}\sum_{i=1}^n x_i$
:::

### Median
::: incremental
- Middle value when data is ordered
- 50th percentile of the data
- More robust to outliers than mean
- For even n, average of two middle values
:::

### Mode
::: incremental
- Most frequently occurring value
- Can have multiple modes
- Only measure of central tendency for categorical data
- Not always meaningful for continuous data
:::

## Visual Comparison of Central Tendency

```{r}
#| label: central-tendency
#| echo: false
#| warning: false
#| fig.height: 4

# Create data with different shapes
set.seed(123)
normal_data <- data.frame(x = rnorm(1000, mean = 0, sd = 1))
skewed_data <- data.frame(x = exp(rnorm(1000, mean = 0, sd = 0.5)))

# Normal distribution plot
p1 <- ggplot(normal_data, aes(x = x)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue", alpha = 0.7) +
  geom_density(color = "darkblue") +
  geom_vline(xintercept = mean(normal_data$x), color = "red", linetype = "dashed") +
  geom_vline(xintercept = median(normal_data$x), color = "green", linetype = "dashed") +
  ggtitle("Symmetric Distribution") +
  theme_minimal() +
  annotate("text", x = mean(normal_data$x), y = 0.1, 
           label = "Mean ≈ Median", angle = 90, vjust = -0.5)

# Skewed distribution plot
p2 <- ggplot(skewed_data, aes(x = x)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightgreen", alpha = 0.7) +
  geom_density(color = "darkgreen") +
  geom_vline(xintercept = mean(skewed_data$x), color = "red", linetype = "dashed") +
  geom_vline(xintercept = median(skewed_data$x), color = "green", linetype = "dashed") +
  ggtitle("Skewed Distribution") +
  theme_minimal() +
  annotate("text", x = mean(skewed_data$x), y = 0.2, 
           label = "Mean", angle = 90, vjust = -0.5) +
  annotate("text", x = median(skewed_data$x), y = 0.2, 
           label = "Median", angle = 90, vjust = -0.5)

# Arrange plots side by side
p1 | p2
```

## Measures of Dispersion

### Variance
::: incremental
- Measures spread of data around the mean
- Average of squared deviations from mean
- Formula: $s^2 = \frac{1}{n-1}\sum_{i=1}^n (x_i - \bar{x})^2$
- Units are squared (harder to interpret)
:::

### Standard Deviation
::: incremental
- Square root of variance
- Same units as original data
- Approximately 68% of data within ±1 SD of mean
- Formula: $s = \sqrt{\frac{1}{n-1}\sum_{i=1}^n (x_i - \bar{x})^2}$
:::

## Visualizing Dispersion

```{r}
#| label: dispersion
#| echo: false
#| warning: false
#| fig.height: 4

# Create datasets with different dispersion
set.seed(456)
narrow_data <- data.frame(x = rnorm(1000, mean = 0, sd = 1))
wide_data <- data.frame(x = rnorm(1000, mean = 0, sd = 2))

# Function to add SD bands
add_sd_bands <- function(p, data, fill_colors) {
  mean_val <- mean(data$x)
  sd_val <- sd(data$x)
  p + 
    geom_rect(aes(xmin = mean_val - sd_val, xmax = mean_val + sd_val, 
                  ymin = -Inf, ymax = Inf), 
              fill = fill_colors[1], alpha = 0.2) +
    geom_rect(aes(xmin = mean_val - 2*sd_val, xmax = mean_val + 2*sd_val, 
                  ymin = -Inf, ymax = Inf), 
              fill = fill_colors[2], alpha = 0.1)
}

# Create base plots
p1 <- ggplot(narrow_data, aes(x = x)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue", color = "black", alpha = 0.7) +
  geom_density(color = "darkblue") +
  ggtitle("Smaller Dispersion (SD = 1)") +
  theme_minimal()

p2 <- ggplot(wide_data, aes(x = x)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightgreen", color = "black", alpha = 0.7) +
  geom_density(color = "darkgreen") +
  ggtitle("Larger Dispersion (SD = 2)") +
  theme_minimal()

# Add SD bands
p1 <- add_sd_bands(p1, narrow_data, c("blue", "lightblue"))
p2 <- add_sd_bands(p2, wide_data, c("green", "lightgreen"))

# Arrange plots
p1 | p2
```




# Thanks for listening! Questions?
This presentation is based on the [SOLES Quarto reveal.js template](https://github.com/usyd-soles-edu/soles-revealjs) and is licensed under a [Creative Commons Attribution 4.0 International License][cc-by]
